Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-02-22 Thesis/01-chap1.Rmd
refine notations

The first model was a cubic linear model with 2 regressors, which can be expressed by: $$\boldsymbol{Y}= 1 + (2-c)\boldsymbol{X} + c\boldsymbol{Z} + a[(2-c)\boldsymbol{X}]^2+a(c\boldsymbol{Z})^2+b[(2-c)\boldsymbol{X}]^3+b(c\boldsymbol{Z})^3+\boldsymbol{\varepsilon},$$ where $c \in (0,2)$, $a \in (-3,3)$, $b \in (-3,3)$, $\boldsymbol{\varepsilon}\overset{iid}{\sim} N(\boldsymbol{0},\sigma^2\boldsymbol{I})$, $\boldsymbol{X}$ and $\boldsymbol{Z}$ are $n\times1$ matrices.

This defines a cubic relationship between $\boldsymbol{Y}$, $\boldsymbol{X}$ and $\boldsymbol{Z}$. Meanwhile, to create nonlinearity defect, the null model fitted by the OLS is: $$\boldsymbol{Y}=\boldsymbol{b}_0+\boldsymbol{b}_1\boldsymbol{X}+\boldsymbol{b}_2\boldsymbol{Z}+\boldsymbol{e}.$$

@@

### Distribution of the Residuals

Let $\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X},\boldsymbol{Z}]$ denotes the set of regressors in matrix form. Using the Frisch–Waugh–Lovell theorem, the residual obtained by the null model is $\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{Y}=\boldsymbol{R}_a(\boldsymbol{X}_a\boldsymbol{\beta}_a+\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}),$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$, $\boldsymbol{\beta}_a=(1,2-c,c)'$, $\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]$ and $\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'$.

Because $\boldsymbol{R}_a\boldsymbol{X}_a=\boldsymbol{0}$, we have $\boldsymbol{e}=\boldsymbol{R}_a(\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}).$ Since $\boldsymbol{\varepsilon} \sim N(0,\sigma^2\boldsymbol{I})$, it follows that $\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a)$.

### Shape of the Residuals

The previous result shows that the expectation of the residuals is clearly a function of $\boldsymbol{X}$ and $\boldsymbol{Z}$. Hence, we would expect to see some patterns in the residual plot. Given the expectation of the residuals, we could plot the expected values of residuals against the observed values. Figure \@ref(fig:cubic1), \@ref(fig:cubic2) and \@ref(fig:cubic3) illustrate the shape of residuals under different parameter settings.

@@

An instance of cubic model class contains methods of simulating data and making residual plot. Method `mod$gen()` returns a data frame containing realizations of $x$, $z$, $y$ and $\varepsilon$ simulated from the model. The number of realizations depends on the integer argument `n`. In addition, if argument `fit_model = TRUE`, a null model will be fitted using the simulated data and residuals and fitted values will be included in the returned data frame.

@@

```{r cubic1, fig.cap = "cubic 1", fig.height=6.3}
library(visage)
library(tidyverse)

@@

expand.grid(c=seq(1),
            a=seq(-1,1,0.5),
            b=seq(-1,1,0.5),
            e_sigma=c(0.5)) %>%
  arrange(a,b,c,e_sigma) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

```{r cubic2, fig.cap = "cubic 2", fig.height=7.875}
expand.grid(c=seq(0.5,1.5,0.25),
            a=seq(-1,1,1),
            b=seq(-1,1,1),
            e_sigma=c(0.5)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

```{r cubic3, fig.cap = "cubic 3", fig.height=7.875}
expand.grid(c=c(1),
            a=seq(-1,1,1),
            b=seq(-1,1,1),
            e_sigma=c(0.25,0.5,1,2)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

result$plots %>% patchwork::wrap_plots(ncol=4)
```

### Effect Size

@@

To obtain a measure of the impact of $a$ and $b$ on the residuals adjusted for variance and covariance, we need to address several properties of the residuals. First, the variance of the residuals $\sigma^2\boldsymbol{R}_a$ is not an identity matrix. This can be fixed by standardizing the residuals by their variance-covariance matrix. Second, the difference between $\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b$ and $\boldsymbol{0}$ could be negative, which is not ideal for comparison. Thus, the magnitude needs to be squared. Third, the measure needs to be a scalar. We could apply a weighted average operator $\boldsymbol{W}$ on the transformed expected residuals to obtain a single numeric value. For simplicity, we set $\boldsymbol{W}=n^{-1}\boldsymbol{1}$. Considering the high time complexity of computing the square root of $\boldsymbol{R}_a$, off-diagonal elements of $\boldsymbol{R}_a$ are set to be zeros. This gives the effect size: $$ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2,$$ where $diag(\boldsymbol{R}_a)$ is the diagonal matrix constructed from the diagnoal element of $\boldsymbol{R}_a$.

@@

```{r effectsize, fig.cap = "Cubic model residual plots under different effect sizes."}
mod <- cubic_model(-1, -1, 1, 0.5)
dat <- mod$gen(500, fit_model = TRUE)
mod$plot(dat, theme = theme_light(base_size = 5), size = 1) +
  ggtitle(glue::glue("Effect size: {round(mod$effect_size(dat), 2)}")) -> p1

@@

###

@@

>>>>

The first model was a cubic linear model with two regressors, which can be expressed by:
\begin{equation} \label{eq:cubic}
\boldsymbol{Y}= 1 + (2-c)\boldsymbol{X} + c\boldsymbol{Z} + a[(2-c)\boldsymbol{X}]^2+a(c\boldsymbol{Z})^2+b[(2-c)\boldsymbol{X}]^3+b(c\boldsymbol{Z})^3+\boldsymbol{\varepsilon},
\end{equation}
where $c \in (0,2)$, $a \in (-3,3)$, $b \in (-3,3)$, $\boldsymbol{\varepsilon}\overset{iid}{\sim} N(\boldsymbol{0},\sigma^2\boldsymbol{I})$, $\boldsymbol{Y}$, $\boldsymbol{X}$ and $\boldsymbol{Z}$ are $n\times1$ matrices.

This defines a cubic relationship between $\boldsymbol{Y}$, $\boldsymbol{X}$ and $\boldsymbol{Z}$. Meanwhile, to create nonlinearity defect, the null model followed the assumptions of the classical normal linear regression model (CNLRM), fitted by OLS is:
\begin{equation} \label{eq:cubic_null}
\boldsymbol{Y}=\beta_0+\beta_1\boldsymbol{X}+\beta_2\boldsymbol{Z}+\boldsymbol{u},
\end{equation}
where $\boldsymbol{u} \sim N(0,\sigma^2_u\boldsymbol{I})$.

@@

\begin{lemma}[Distribution of residuals produced by the cubic model] \label{lemma:cubic}
Given the data generating process in Equation (\ref{eq:cubic}), and null model in Equation (\ref{eq:cubic_null}). Let $\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X},\boldsymbol{Z}]$ denotes the set of regressors in matrix form. Then, the residuals obtained from the null model are $$\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$, $\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]$ and $\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'$.
\end{lemma}

\begin{proof}
Using the Frisch–Waugh–Lovell theorem, the residuals obtained by the null model are $$\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{Y}=\boldsymbol{R}_a(\boldsymbol{X}_a\boldsymbol{\beta}_a+\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$, $\boldsymbol{\beta}_a=(1,2-c,c)'$, $\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]$ and $\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'$.

Because $\boldsymbol{R}_a\boldsymbol{X}_a=\boldsymbol{0}$, we have $\boldsymbol{e}=\boldsymbol{R}_a(\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}).$ Since $\boldsymbol{\varepsilon} \sim N(0,\sigma^2\boldsymbol{I})$, it follows that $\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a)$. 
\end{proof}

Lemma \@ref(lemma:cubic) shows that the expectation of the residuals is clearly a function of $\boldsymbol{X}$ and $\boldsymbol{Z}$. Hence, we would expect to see some patterns in the residual plot. Given the expectation of the residuals, we could plot the expected values of residuals against the observed values. Figure \@ref(fig:cubic1), \@ref(fig:cubic2) and \@ref(fig:cubic3) illustrate the shape of residuals under different parameter settings.

@@

An instance of cubic model class contains methods of simulating data and making residual plot. Method `mod$gen()` returns a data frame containing realizations of $X$, $Z$, $Y$ and $\varepsilon$ simulated from the model. The number of realizations depends on the integer argument `n`. In addition, if argument `fit_model = TRUE`, a null model will be fitted using the simulated data and residuals and fitted values will be included in the returned data frame.

@@

```{r cubic1, fig.cap = "cubic 1"}
library(visage)
library(tidyverse)

@@

expand.grid(c=seq(1),
            a=seq(-1,1,1),
            b=seq(-1,1,1),
            e_sigma=c(0.5)) %>%
  arrange(a,b,c,e_sigma) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

result$plots %>% patchwork::wrap_plots(ncol=3)
```

```{r cubic2, fig.cap = "cubic 2", fig.height=1.9}
expand.grid(c=c(0,0.5,1,1.5,2),
            a=c(-1),
            b=c(1),
            e_sigma=c(0.5)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

```{r cubic3, fig.cap = "cubic 3", fig.height=1.9}
expand.grid(c=c(1),
            a=c(-1),
            b=c(1),
            e_sigma=c(0.25,0.5,1,2,4)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

@@

To obtain a measure of the impact of $a$ and $b$ on the residuals adjusted for variance and covariance, we need to address several properties of the residuals. First, the variance of the residuals $\sigma^2\boldsymbol{R}_a$ is not an identity matrix. This can be fixed by standardizing the residuals by their variance-covariance matrix. Second, the difference between $\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b$ and $\boldsymbol{0}$ could be negative, which is not ideal for comparison. Thus, the magnitude needs to be squared. Third, the measure needs to be a scalar. We could apply a weighted average operator $\boldsymbol{W}$ on the transformed expected residuals to obtain a single numeric value. For simplicity, we set $\boldsymbol{W}=n^{-1}\boldsymbol{1}$. Considering the high time complexity of computing the square root of $\boldsymbol{R}_a$, off-diagonal elements of $\boldsymbol{R}_a$ are set to be zeros. This gives the effect size: $$ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2,$$ where $diag(\boldsymbol{R}_a)$ is the diagonal matrix constructed from the diagonal elements of $\boldsymbol{R}_a$.

@@

```{r effectsize, fig.cap = "Cubic model residual plots under different effect sizes. The larger the effect size, the stronger the signal."}
mod <- cubic_model(-1, -1, 1, 0.5)
dat <- mod$gen(500, fit_model = TRUE)
mod$plot(dat, theme = theme_light(base_size = 5), size = 1) +
  ggtitle(glue::glue("Effect size: {round(mod$effect_size(dat), 2)}")) -> p1

@@

Under the cubic model setting, there is an exact conventional test for testing the nonlinearity defect, which is F-test. For F-test, the null hypothesis is $H_0:a=b=0$, and the alternative hypothesis is $H_1:\text{at least one of them} \neq 0$. During the simulation of the lineup data, the F-statistic and the p-value will be recorded for comparison between the power of conventional test and visual test.

@@

Another model used in the experiments was a heteroskedasticity model with one regressor, which can be expressed by: 
\begin{equation} \label{eq:heter}
Y_i = 1 + X_i + \varepsilon_i, ~i = 1,...,n,
\end{equation}
where $a \in \{-1,0,1\}$, $b\in (0,32)$ and $\varepsilon_i \overset{iid}{\sim} N(0,1+b(2-|a|)(X_i-a)^2)$.

To create heteroskedasticity defect, OLS was used to fit the null model:
\begin{equation} \label{eq:heter_null}
\boldsymbol{Y}=\beta_0+\beta_1\boldsymbol{X}+\boldsymbol{u},
\end{equation}
where $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2_u\boldsymbol{I})$.

In this case, estimators of $\beta_0$ and $\beta_1$ are unbiased, but the error term has non-constant variance.

\begin{lemma}[Distribution of residuals produced by the heteroskedasticity model] \label{lemma:heter}
Given the data generating process in Equation (\ref{eq:heter}) and null model in Equation (\ref{eq:heter_null}). Let $\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X}]$ denotes the set of regressors in matrix form. The residuals obtained from the null model are $$\boldsymbol{e} \sim N(\boldsymbol{0}, \boldsymbol{R}_a\boldsymbol{V}),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ and $\boldsymbol{V}$ is a diagonal matrix with $V_{ii}=1+b(2-|a|)(X_i  -  a)^2$, $i = 1,...,n$. 
\end{lemma}

\begin{proof}
Using the Frisch–Waugh–Lovell theorem, the residuals obtained by the null model are $\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{Y}=\boldsymbol{R}_a(\boldsymbol{X}_a\boldsymbol{\beta}_a+\boldsymbol{\varepsilon}),$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ and $\boldsymbol{\beta}_a=(1,1)'$.

Because $\boldsymbol{R}_a\boldsymbol{X}_a=\boldsymbol{0}$, we have $\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{\varepsilon}$. Hence, the residuals $\boldsymbol{e}$ follow $N(\boldsymbol{0}, \boldsymbol{R}_a\boldsymbol{V})$, where $\boldsymbol{V}$ is a diagonal matrix with $V_{ii}=1+b(2-|a|)(X_i  -  a)^2$, $i = 1,...,n$. 
\end{proof}

Lemma \@ref(lemma:heter) shows

This shows that the variance of the residuals is clearly a function of X and Z. Hence, we
would expect to see some patterns in the residual plot.

## Distribution of regressors

### 2022-02-22 Thesis/_book/thesis.tex
refine notations

\begin{document}

@@

The first model was a cubic linear model with 2 regressors, which can be expressed by: \[\boldsymbol{Y}= 1 + (2-c)\boldsymbol{X} + c\boldsymbol{Z} + a[(2-c)\boldsymbol{X}]^2+a(c\boldsymbol{Z})^2+b[(2-c)\boldsymbol{X}]^3+b(c\boldsymbol{Z})^3+\boldsymbol{\varepsilon},\] where \(c \in (0,2)\), \(a \in (-3,3)\), \(b \in (-3,3)\), \(\boldsymbol{\varepsilon}\overset{iid}{\sim} N(\boldsymbol{0},\sigma^2\boldsymbol{I})\), \(\boldsymbol{X}\) and \(\boldsymbol{Z}\) are \(n\times1\) matrices.

This defines a cubic relationship between \(\boldsymbol{Y}\), \(\boldsymbol{X}\) and \(\boldsymbol{Z}\). Meanwhile, to create nonlinearity defect, the null model fitted by the OLS is: \[\boldsymbol{Y}=\boldsymbol{b}_0+\boldsymbol{b}_1\boldsymbol{X}+\boldsymbol{b}_2\boldsymbol{Z}+\boldsymbol{e}.\]

@@

\hypertarget{distribution-of-the-residuals}{%
\subsection{Distribution of the Residuals}\label{distribution-of-the-residuals}}

Let \(\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X},\boldsymbol{Z}]\) denotes the set of regressors in matrix form. Using the Frisch--Waugh--Lovell theorem, the residual obtained by the null model is \(\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{Y}=\boldsymbol{R}_a(\boldsymbol{X}_a\boldsymbol{\beta}_a+\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}),\) where \(\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'\), \(\boldsymbol{\beta}_a=(1,2-c,c)'\), \(\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]\) and \(\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'\).

Because \(\boldsymbol{R}_a\boldsymbol{X}_a=\boldsymbol{0}\), we have \(\boldsymbol{e}=\boldsymbol{R}_a(\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}).\) Since \(\boldsymbol{\varepsilon} \sim N(0,\sigma^2\boldsymbol{I})\), it follows that \(\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a)\).

\hypertarget{shape-of-the-residuals}{%
\subsection{Shape of the Residuals}\label{shape-of-the-residuals}}

The previous result shows that the expectation of the residuals is clearly a function of \(\boldsymbol{X}\) and \(\boldsymbol{Z}\). Hence, we would expect to see some patterns in the residual plot. Given the expectation of the residuals, we could plot the expected values of residuals against the observed values. Figure \ref{fig:cubic1}, \ref{fig:cubic2} and \ref{fig:cubic3} illustrate the shape of residuals under different parameter settings.

@@

An instance of cubic model class contains methods of simulating data and making residual plot. Method \texttt{mod\$gen()} returns a data frame containing realizations of \(x\), \(z\), \(y\) and \(\varepsilon\) simulated from the model. The number of realizations depends on the integer argument \texttt{n}. In addition, if argument \texttt{fit\_model\ =\ TRUE}, a null model will be fitted using the simulated data and residuals and fitted values will be included in the returned data frame.

@@

\hypertarget{effect-size}{%
\subsection{Effect Size}\label{effect-size}}

@@

To obtain a measure of the impact of \(a\) and \(b\) on the residuals adjusted for variance and covariance, we need to address several properties of the residuals. First, the variance of the residuals \(\sigma^2\boldsymbol{R}_a\) is not an identity matrix. This can be fixed by standardizing the residuals by their variance-covariance matrix. Second, the difference between \(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b\) and \(\boldsymbol{0}\) could be negative, which is not ideal for comparison. Thus, the magnitude needs to be squared. Third, the measure needs to be a scalar. We could apply a weighted average operator \(\boldsymbol{W}\) on the transformed expected residuals to obtain a single numeric value. For simplicity, we set \(\boldsymbol{W}=n^{-1}\boldsymbol{1}\). Considering the high time complexity of computing the square root of \(\boldsymbol{R}_a\), off-diagonal elements of \(\boldsymbol{R}_a\) are set to be zeros. This gives the effect size: \[ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2,\] where \(diag(\boldsymbol{R}_a)\) is the diagonal matrix constructed from the diagnoal element of \(\boldsymbol{R}_a\).

@@

\begin{figure}
\centering
\includegraphics{thesis_files/figure-latex/effectsize-1.pdf}
\caption{\label{fig:effectsize}Cubic model residual plots under different effect sizes.}
\end{figure}

\hypertarget{section}{%
\subsection{}\label{section}}

@@

>>>>

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\newtheorem{conjecture}{Conjecture}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\theoremstyle{definition}
\newtheorem{example}{Example}[chapter]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[chapter]
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}[chapter]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

The first model was a cubic linear model with two regressors, which can be expressed by:
\begin{equation} \label{eq:cubic}
\boldsymbol{Y}= 1 + (2-c)\boldsymbol{X} + c\boldsymbol{Z} + a[(2-c)\boldsymbol{X}]^2+a(c\boldsymbol{Z})^2+b[(2-c)\boldsymbol{X}]^3+b(c\boldsymbol{Z})^3+\boldsymbol{\varepsilon},
\end{equation}
where \(c \in (0,2)\), \(a \in (-3,3)\), \(b \in (-3,3)\), \(\boldsymbol{\varepsilon}\overset{iid}{\sim} N(\boldsymbol{0},\sigma^2\boldsymbol{I})\), \(\boldsymbol{Y}\), \(\boldsymbol{X}\) and \(\boldsymbol{Z}\) are \(n\times1\) matrices.

This defines a cubic relationship between \(\boldsymbol{Y}\), \(\boldsymbol{X}\) and \(\boldsymbol{Z}\). Meanwhile, to create nonlinearity defect, the null model followed the assumptions of the classical normal linear regression model (CNLRM), fitted by OLS is:
\begin{equation} \label{eq:cubic_null}
\boldsymbol{Y}=\beta_0+\beta_1\boldsymbol{X}+\beta_2\boldsymbol{Z}+\boldsymbol{u},
\end{equation}
where \(\boldsymbol{u} \sim N(0,\sigma^2_u\boldsymbol{I})\).

@@

\begin{lemma}[Distribution of residuals produced by the cubic model] \label{lemma:cubic}
Given the data generating process in Equation (\ref{eq:cubic}), and null model in Equation (\ref{eq:cubic_null}). Let $\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X},\boldsymbol{Z}]$ denotes the set of regressors in matrix form. Then, the residuals obtained from the null model are $$\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$, $\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]$ and $\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'$.
\end{lemma}

\begin{proof}
Using the Frisch–Waugh–Lovell theorem, the residuals obtained by the null model are $$\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{Y}=\boldsymbol{R}_a(\boldsymbol{X}_a\boldsymbol{\beta}_a+\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$, $\boldsymbol{\beta}_a=(1,2-c,c)'$, $\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]$ and $\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'$.

Because $\boldsymbol{R}_a\boldsymbol{X}_a=\boldsymbol{0}$, we have $\boldsymbol{e}=\boldsymbol{R}_a(\boldsymbol{X}_b\boldsymbol{\beta}_b+\boldsymbol{\varepsilon}).$ Since $\boldsymbol{\varepsilon} \sim N(0,\sigma^2\boldsymbol{I})$, it follows that $\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a)$. 
\end{proof}

Lemma \ref{lemma:cubic} shows that the expectation of the residuals is clearly a function of \(\boldsymbol{X}\) and \(\boldsymbol{Z}\). Hence, we would expect to see some patterns in the residual plot. Given the expectation of the residuals, we could plot the expected values of residuals against the observed values. Figure \ref{fig:cubic1}, \ref{fig:cubic2} and \ref{fig:cubic3} illustrate the shape of residuals under different parameter settings.

@@

An instance of cubic model class contains methods of simulating data and making residual plot. Method \texttt{mod\$gen()} returns a data frame containing realizations of \(X\), \(Z\), \(Y\) and \(\varepsilon\) simulated from the model. The number of realizations depends on the integer argument \texttt{n}. In addition, if argument \texttt{fit\_model\ =\ TRUE}, a null model will be fitted using the simulated data and residuals and fitted values will be included in the returned data frame.

@@

@@

To obtain a measure of the impact of \(a\) and \(b\) on the residuals adjusted for variance and covariance, we need to address several properties of the residuals. First, the variance of the residuals \(\sigma^2\boldsymbol{R}_a\) is not an identity matrix. This can be fixed by standardizing the residuals by their variance-covariance matrix. Second, the difference between \(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b\) and \(\boldsymbol{0}\) could be negative, which is not ideal for comparison. Thus, the magnitude needs to be squared. Third, the measure needs to be a scalar. We could apply a weighted average operator \(\boldsymbol{W}\) on the transformed expected residuals to obtain a single numeric value. For simplicity, we set \(\boldsymbol{W}=n^{-1}\boldsymbol{1}\). Considering the high time complexity of computing the square root of \(\boldsymbol{R}_a\), off-diagonal elements of \(\boldsymbol{R}_a\) are set to be zeros. This gives the effect size: \[ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2,\] where \(diag(\boldsymbol{R}_a)\) is the diagonal matrix constructed from the diagonal elements of \(\boldsymbol{R}_a\).

@@

\begin{figure}
\centering
\includegraphics{thesis_files/figure-latex/effectsize-1.pdf}
\caption{\label{fig:effectsize}Cubic model residual plots under different effect sizes. The larger the effect size, the stronger the signal.}
\end{figure}

Under the cubic model setting, there is an exact conventional test for testing the nonlinearity defect, which is F-test. For F-test, the null hypothesis is \(H_0:a=b=0\), and the alternative hypothesis is \(H_1:\text{at least one of them} \neq 0\). During the simulation of the lineup data, the F-statistic and the p-value will be recorded for comparison between the power of conventional test and visual test.

@@

Another model used in the experiments was a heteroskedasticity model with one regressor, which can be expressed by:
\begin{equation} \label{eq:heter}
Y_i = 1 + X_i + \varepsilon_i, ~i = 1,...,n,
\end{equation}
where \(a \in \{-1,0,1\}\), \(b\in (0,32)\) and \(\varepsilon_i \overset{iid}{\sim} N(0,1+b(2-|a|)(X_i-a)^2)\).

To create heteroskedasticity defect, OLS was used to fit the null model:
\begin{equation} \label{eq:heter_null}
\boldsymbol{Y}=\beta_0+\beta_1\boldsymbol{X}+\boldsymbol{u},
\end{equation}
where \(\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2_u\boldsymbol{I})\).

In this case, estimators of \(\beta_0\) and \(\beta_1\) are unbiased, but the error term has non-constant variance.

\begin{lemma}[Distribution of residuals produced by the heteroskedasticity model] \label{lemma:heter}
Given the data generating process in Equation (\ref{eq:heter}) and null model in Equation (\ref{eq:heter_null}). Let $\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X}]$ denotes the set of regressors in matrix form. The residuals obtained from the null model are $$\boldsymbol{e} \sim N(\boldsymbol{0}, \boldsymbol{R}_a\boldsymbol{V}),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ and $\boldsymbol{V}$ is a diagonal matrix with $V_{ii}=1+b(2-|a|)(X_i  -  a)^2$, $i = 1,...,n$. 
\end{lemma}

\begin{proof}
Using the Frisch–Waugh–Lovell theorem, the residuals obtained by the null model are $\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{Y}=\boldsymbol{R}_a(\boldsymbol{X}_a\boldsymbol{\beta}_a+\boldsymbol{\varepsilon}),$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ and $\boldsymbol{\beta}_a=(1,1)'$.

Because $\boldsymbol{R}_a\boldsymbol{X}_a=\boldsymbol{0}$, we have $\boldsymbol{e}=\boldsymbol{R}_a\boldsymbol{\varepsilon}$. Hence, the residuals $\boldsymbol{e}$ follow $N(\boldsymbol{0}, \boldsymbol{R}_a\boldsymbol{V})$, where $\boldsymbol{V}$ is a diagonal matrix with $V_{ii}=1+b(2-|a|)(X_i  -  a)^2$, $i = 1,...,n$. 
\end{proof}

Lemma \ref{lemma:heter} shows

This shows that the variance of the residuals is clearly a function of X and Z. Hence, we
would expect to see some patterns in the residual plot.

\hypertarget{distribution-of-regressors}{%
\section{Distribution of regressors}\label{distribution-of-regressors}}
